---
title: "Home"
---

The 2008 financial crisis triggered a global recession, disrupting trade and reducing demand across sectors. Its impact on U.S. live animal imports may reveal how major economic shocks can cause lasting changes in trade behavior.


### Question


> Did the 2008 global financial crisis cause a long-term shift in U.S. live animal import behavior?

###

Understanding this shift can reveal how economic shocks reshape trade in essential goods and help policymakers, economists, and industry stakeholders anticipate and adapt to future disruptions.

We analyzed U.S. live animal imports from 2000 to 2016 using the Global Commodity Trade Statistics dataset. Post-2008 trends show shifts in import volume and value, suggesting the crisis may have triggered lasting changes in trade behavior.

```{r}
library(tidyverse)
library(tidymodels)
library(primer.data)
library(knitr)
library(kableExtra)
library(broom)
library(gt)
library(stringr)
library(fixest)

gct <- read_csv("commodity_trade_statistics_data.csv")
```

```{r}
library(tidyverse)
library(scales)

us_live_all <- gct %>%
  filter(
    country_or_area == "USA",
    category == "01_live_animals",
    flow == "Import",
    year >= 2000, year <= 2016,
    !is.na(trade_usd)
  ) %>%
  group_by(year) %>%
  summarize(total_trade = sum(trade_usd, na.rm = TRUE), .groups = "drop") %>%
  mutate(year_factor = as.factor(year))

ggplot(us_live_all, aes(x = year_factor, y = total_trade / 1e9, group = 1)) +
  geom_area(fill = "#CAC4A0", alpha = 0.8) +                   # light tan fill
  geom_line(color = "#6D624A", linewidth = 1.2) +              # dark brown outline line
  coord_polar(start = -pi/16) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "$", suffix = "B")) +
  labs(
    title = "U.S. Live Animal Imports (2000–2016)",
    x = NULL,
    y = "Trade Value (Billions USD)"
  ) +
  theme_minimal(base_family = "Segoe UI") +
  theme(
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.major.y = element_line(color = "gray80"),
    panel.grid.minor = element_blank(),
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 16)
  )

```

The plot reveals fluctuations in U.S. live animal imports over the years, indicating periods of increase and decline in trade value between 2000 and 2016.
