---
title: "Globe"
format:
  html:
    theme: cosmo
    css: styles.css
jupyter: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(tidymodels)
library(primer.data)
library(knitr)
library(kableExtra)
library(broom)
library(gt)
library(stringr)
library(fixest)

gct <- read_csv("commodity_trade_statistics_data.csv")
```

```{r}
library(tidyverse)
library(plotly)
library(rnaturalearth)
library(rnaturalearthdata)
library(countrycode)

# Pre-process data
gct_filtered <- gct %>%
  filter(flow == "Import", !is.na(trade_usd)) %>%
  mutate(period = if_else(year <= 2008, "Pre-2008", "Post-2008"))

# Aggregate by country and period
gct_map <- gct_filtered %>%
  group_by(country_or_area, period) %>%
  summarize(total_trade = sum(trade_usd, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    trade_label = case_when(
      total_trade >= 1e12 ~ paste0("$", round(total_trade / 1e12, 1), "T"),
      total_trade >= 1e9 ~ paste0("$", round(total_trade / 1e9, 1), "B"),
      total_trade >= 1e6 ~ paste0("$", round(total_trade / 1e6, 1), "M"),
      total_trade > 0 ~ paste0("Small value ($", round(total_trade, 0), ")"),
      total_trade == 0 ~ "No trade reported",
      TRUE ~ NA_character_
    ),
    iso_a3 = countrycode(country_or_area, "country.name", "iso3c")
  )

# Prepare world map
world <- ne_countries(scale = "medium", returnclass = "sf")

# Join for both periods
world_data <- world %>%
  select(iso_a3, name) %>%
  right_join(gct_map, by = "iso_a3") %>%
  mutate(
    z = ifelse(is.na(total_trade), -1, total_trade),
    hover_text = ifelse(is.na(total_trade),
                        paste(name, "<br>No data"),
                        paste(name, "<br>", trade_label))
  )

# Define color scale
color_scale <- list(
  list(0, "rgb(200,200,200)"),
  list(0.000001, "#f5f0e1"),
  list(0.33, "#c2b280"),
  list(0.66, "#8c6d31"),
  list(1, "#5c4422")
)

# Separate frames for pre/post
frames <- split(world_data, world_data$period)

# Create initial plot (Pre-2008)
p <- plot_ly() %>%
  add_trace(
    type = "choropleth",
    locations = frames[["Pre-2008"]]$iso_a3,
    z = frames[["Pre-2008"]]$z,
    text = frames[["Pre-2008"]]$hover_text,
    colorscale = color_scale,
    zmin = -1,
    zmax = max(world_data$z, na.rm = TRUE),
    hoverinfo = "text",
    marker = list(line = list(width = 0)),
    showscale = TRUE,
    colorbar = list(title = "Trade Value", tickprefix = "$")
  ) %>%
  layout(
    geo = list(
      projection = list(type = "orthographic"),
      showland = TRUE,
      showcountries = TRUE,
      landcolor = "gray90",
      oceancolor = "white",
      countrycolor = "gray20",
      bgcolor = "white",
      showframe = FALSE,
      showcoastlines = FALSE
    ),
    updatemenus = list(
      list(
        type = "buttons",
        direction = "right",
        x = 0.5,
        y = 1.05,
        showactive = TRUE,
        buttons = list(
          list(
            label = "Pre-2008",
            method = "restyle",
            args = list(list(
              locations = list(frames[["Pre-2008"]]$iso_a3),
              z = list(frames[["Pre-2008"]]$z),
              text = list(frames[["Pre-2008"]]$hover_text)
            ))
          ),
          list(
            label = "Post-2008",
            method = "restyle",
            args = list(list(
              locations = list(frames[["Post-2008"]]$iso_a3),
              z = list(frames[["Post-2008"]]$z),
              text = list(frames[["Post-2008"]]$hover_text)
            ))
          )
        )
      )
    ),
    annotations = list(
      list(
        text = "Note: Yellow countries have missing trade data",
        x = 0.5,
        y = -0.1,
        showarrow = FALSE,
        xref = "paper",
        yref = "paper",
        font = list(size = 12, color = "gray20")
      )
    )
  )

p

```

This globe shows the total recorded trade value for each country, based on all years and commodities in the dataset. Each country reflects its cumulative imports and exports in U.S. dollars, representing its overall trade activity during the observed period.

Countries without data are shown in gray. These include:

Old countries or dissolved regions (e.g., Serbia and Montenegro),

Trade zones or regional groupings (e.g., EU-28, Other Asia, nes),

Or countries with incomplete or missing records.

The globe highlights where trade was most heavily reported and where gaps exist in the data. It gives a high-level view of how global trade is distributed and where it’s missing.
